---
title: "Problem Set 8"
subtitle: "Due date: 20 November"
format: 
  html:
    self-contained: true
toc: true
editor: visual
execute: 
  echo: true
---

Please upload your completed assignment to the ELMs course site (under the assignments menu). Remember to include an annotated script file for all work with R and show your math for all other problems (if applicable, or necessary). Please also upload your completed assignment to the Github repository that you have shared with us. *We should be able to run your script with no errors.*

**Total points: 30**

## Question 1

*Points: 5*

::: {.callout-tip}
5/5
:::

For the following regression equation, $\hat{Y} = 8.5 + 6x + \epsilon$, the standard error for $\beta_0$ is 2.5, the standard error for $\beta_1$ is 3.5, and the sample size is 2000. Find the t-statistic, 95% confidence interval, and p-value (using a two-tailed test) for $\beta_1$.

Is $\beta_1$ statistically significant at the 0.05-level with a two-tailed test? Why or why not?

t-statistic:

$t = \frac{6-0}{3.5} = 1.71$

The 95% CI:

Upper bound: 6 + 1.961 x 3.5 = 12.86

Lower bound: 6 -1.961 x 3.5 = -0.86

The 95% confidence interval for $\beta_1$​ is approximately from -0.86 to 12.86. The interval suggests that, with 95% confidence, we can say the true value of $\beta_1$​​ lies somewhere within this range.

P-value:\
With a t-statistic of 1.71 and 1998 degrees of freedom, we get a p-value of .087

β1​ is not statistically significant at the 0.05 level based on the t-statistic, p-value, and confidence interval, meaning we don't have enough evidence to say that $\beta_1$​ is different from 0.

## Question 2

*Points: 5*

::: {.callout-tip}
5/5
:::

Suppose you estimate an OLS regression and retrieve a $R^2$ value of 0.45. If the Total Sum of Squares (TSS) from that regression equals 4,700, what is the value for the Residual Sum of Squares (RSS)?

$RSS = 4700\times(1-0.45)= 2585$

This tells us that the total variance in the DV, the models residuals account for 2585 units.

## Question 3

*Points: 5*

::: {.callout-tip}
5/5
:::

Suppose you estimate a bivariate regression with a sample size of 102 and obtain a regression coefficient ($\beta_1$) of 5.0. What is the largest standard error that $\beta_1$ could have and still be statistically significant (i.e., reject the null hypothesis of no relationship) at the 0.05 level with a one-tailed test?

$SE(β1​) = \frac{5.0}{1.66} = 3.01$

The largest standard error for $\beta_1$ that still allows for statistical significance at the 0.05 level about 3.01. This means if the actual SE of $\beta_1$ ​ is less than or equal to 3.01, the coefficient is statistically significant at the 5% level, which will reject the null.

## Question 4

*Points: 5*

::: {.callout-tip}
5/5
:::

Using the `states` dataset from the `poliscidata` package, produce a scatterplot of the variables `romney2012` and `hispanic10` (with `romney2012` as the dependent variable on the y-axis). Fit a regression line to the scatterplot. Describe the scatterplot and include a copy of it. Note any suspected outliers, if any (a visual inspection will suffice for this question).

::: callout-note
The variable `romney2012` measures the percentage of the state's vote that Mitt Romney received in the 2012 presidential election, and `hispanic10` indicates the percentage of the state's population that identified as Hispanic in 2010.
:::

```{r}
library(poliscidata)
library(modelsummary)
library(tidyverse)
library(ggplot2)
ggplot(states, aes(x = hispanic10, y = romney2012))+
  geom_point()+
  geom_smooth(method = lm, se = F)+
  theme_classic()

model <- lm(romney2012 ~ hispanic10, data = states)
modelsummary(model,
             statistic = NULL,
             gof_map = "nobs")

cor(states$romney2012, states$hispanic10)

```

Strength: negative

Direction:

This scatterplot shows a negative relationship between the percentages of the votes Mitt Romney recieve in a state and the percentage of the hispanic population. With a correlation coefficient of -0.19.

## Question 5

*Points: 10*

::: {.callout-tip}
8/10
:::

Estimate a bivariate regression with `romney2012` as the dependent variable and `hispanic10` as the independent variable and report the results in a professionally formatted table. In as much detail as possible, describe (and interpret) the regression results.

```{r}
model <- lm(romney2012 ~ hispanic10, data = states)
summary(model)
```

Our regression result shows that the intercept is 51.87, with a 95% confident interval of between 47.65 and 56.09, this means that if the percentage of the hispanic population were zero the predicted outcome would be 51.87. 

::: {.callout-tip}
Make sure to always say that this is the effect *on average*.
:::

The coefficint in out model shows -0.196 meaning that for every one percentage point increase in the Hispanic population within a state, the dependent variable is expected to decrease by 0.196 percentage points, holding all else constant. 

::: {.callout-tip}
Again, *on average*.
:::

Our model shows an R-squared of 0.037, which means that approximately 3.7% of the variance in the dependent variable is explained by the x variable. This is a very low value, indicating that the model does not explain much of the variance. Similarly, the adjusted R-squared value is 0.017,which further confirms that the model explains very little of the variance in the dependent variable. Finally, the p-value associated with the coefficient is 0.182, since it is greater than 0.05, the statistical evidence is not strong enough to reject the null at the 5% significance level. Therefore, we would not consider the relationship between the percentage of the Hispanic population and the dependent variable to be statistically significant based on this model.
